---
title: "Reproducible Research: Peer Assessment 1"
author: "Vidya Sankaran"
date: "December 8, 2017"
output: 
      html_document:
         keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Loading and preprocessing the data
Including Required library files and downloading the zip file. Unziping the same and read the dataset. Inspecting the structure of the variable and identify the contents. 
Following transformations are done:

- Month - calculation from the date given


```{r Load and Preprocessing}
library(dplyr)
library(readr)
library(ggplot2)
library(tidyr)
library(lubridate)
library(stringr)
#library(Hmisc)

if (file.exists("activity.zip") == FALSE) {
    download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", destfile = "activity.zip")  
    unzip("activity.zip")
}

mydf <- read.csv("activity.csv", header = TRUE)
str(mydf)
dim(mydf)
head(mydf)
mean(is.na(mydf$steps))

z<- ymd(mydf$date, tz="GMT")

mydf <- mydf %>% mutate(month_num = month(z)) %>%
        mutate(daycount = ifelse(month_num==10, day(z), day(z)+31))
tail(mydf, n=10)

```

Finding:
61 days which is Oct 31 days, and Nov 30 days the readings were taken.


## What is mean total number of steps taken per day?
Following transformations done for answering this question
- Daycount: How many days these readings are taken 
- Steps per day: Total no of steps taken in every day excluding the NA values

Plot:
    A histogram which represents the no of steps taken each day 


```{r Total No of steps and Median and Mean of steps taken each day}

mydf <- mydf %>% group_by(daycount) %>% mutate(steps_per_day = sum(steps, na.rm=TRUE)) 

mydf_steps <- mydf %>% group_by(daycount, date) %>% summarise(steps_per_day1=sum(steps,na.rm=TRUE))
mydf_steps

hist(x=mydf_steps$steps_per_day1, main="Total No of Steps Taken Each Day", col="blue") 
abline(v=mean(mydf_steps$steps_per_day1), col="red", lwd=2)
abline(v=median(mydf_steps$steps_per_day1), col="black", lwd=2)

summary(mydf_steps$steps_per_day1)

mydf

```

 Findings:
 For about 13 days, the steps value is zero. This may be due to NA or due to some
 reason the person not measured it. 
 The person takes about average of 9354 steps in 61 days and the median is 10395 steps


## What is the average daily activity pattern?

For this question following transformation done
-  Padding of time : The readings are taken in 5 minute interval in a day, so 
converted the time to proper 24 hour time format for Hours and                       minutes, so, padded the data with 0s for the previous blank                          lines. Stringr library included for this purpose

-  POSIXct datetime : conversion of date and time values to date time mode
-  average steps taken per day : by grouping the datetime and daycount, we 
summaraise the mean steps taken per day   

- Steps per interval:  Steps for that interval including all 61 days in the dataset
            are taken and added by excluding the NA values.

- five_min_interval:  The pad time is converted to numerical factor number, says, 0000 to 1, 0005 to 2 and 23:55 as 288  This will be useful for showing graphicaly

```{r  5-minute interval containing max no of steps}


mydf <- mydf %>% mutate(pad_time= str_pad(as.character(interval), 4, pad="0")) %>%
    mutate(datetime =as.POSIXct(paste(date, pad_time), tz="GMT"))

head(mydf, n=3)
mean_mydf <- mean(mydf$steps, na.rm=TRUE)

mydf2 <- mydf %>% group_by(daycount, datetime ) %>% 
    select(daycount, datetime,  steps_per_day) %>% 
    summarise (avg_steps_day  = mean(steps_per_day)  )

mydf2

mydf1 <- mydf %>% group_by(pad_time)  %>% 
    summarise(avg_steps_per_interval = mean(steps, na.rm=TRUE))%>% 
    select(pad_time,  avg_steps_per_interval ) 


mydf1

mydf1 <- mydf1 %>% mutate(five_min_interval = 1:nrow(mydf1) )

mydf1

tmp <- max(mydf1$avg_steps_per_interval)    
xinte <- mydf1$pad_time[mydf1$avg_steps_per_interval==max(mydf1$avg_steps_per_interval) ]  
xinte

xinte <- mydf1$five_min_interval[mydf1$avg_steps_per_interval==max(mydf1$avg_steps_per_interval)]


summary(mydf1$pad_time)



ggplot(mydf1, aes(five_min_interval, avg_steps_per_interval))+ 
    geom_line( col="blue")+  
    ggtitle("Steps per day with respect to 5 minute time interval") +
    geom_vline(xintercept = xinte, col="brown")+
    xlab(" 5 minute interval as numeric values from 0000 to 1 and  23:55 as 288")
    

head(mydf, n=1)

```


Findings:

The 104th interval, which is at 0835, the maximum no of steps recorded.


## Imputing missing values
strategy for missing data:

_ When checked all the days are weekdays for the imputing.
- We can impute mean of the steps per interval instead of NA values

```{r imputting missing data}
library(Hmisc)
temp <- which(is.na(mydf$steps))
wday(mydf$datetime[1], label=TRUE)

#One and seven are sunday and saturday
mydf_imp <- mydf %>% group_by(pad_time) %>% 
    select (datetime, daycount, pad_time, steps) %>% 
    filter(is.na(steps)) 

mydf_imp



mydf
mydf$steps <-impute(mydf$steps, mean)
mydf

mydf <- mydf %>% group_by(daycount) %>% 
    mutate(weekday = wday(datetime)) %>% 
    mutate(steps_per_day= sum(steps))
mydf

hist(x=mydf$steps_per_day, main="Total No of Steps Taken Each Day", col="blue")
abline(v=mean(mydf$steps_per_day), col="red", lwd=2)
abline(v=median(mydf$steps_per_day), col="black", lwd=2)

summary(mydf$steps_per_day)



```

Findings:

By imputing the average of steps taken per interval, the mean and median of the steps per day has increased and now they are at same value.

## Are there differences in activity patterns between weekdays and weekends?


```{r Weekdays Vs Weekends}

mydf

mydf <- mydf %>% group_by(pad_time) %>% 
    mutate(weekend= ifelse(weekday==1 | weekday==7, TRUE, FALSE))

mydf


#mydf_wk_day <- mydf %>% group_by(daycount, date) %>%
#    select(date, daycount, datetime, steps_per_day, weekend) %>% 
#    filter (weekend ==FALSE)

mydf_wk_day <- mydf %>% group_by(pad_time) %>%
    select(pad_time, steps, weekend) %>% 
    filter (weekend ==FALSE) %>%
    summarise(steps_per_interval = mean(steps))%>%
    select(pad_time, steps_per_interval)
    
  
mydf_wk_day


#mydf_wk_end <- mydf %>% group_by(daycount, date) %>%
#    select(date, daycount, datetime, steps_per_day, weekend) %>% 
#    filter (weekend ==TRUE)

mydf_wk_end <- mydf %>% group_by(pad_time) %>%
    select(pad_time, steps, weekend) %>% 
    filter (weekend ==TRUE) %>%
    summarise(steps_per_interval = mean(steps))%>%
    select(pad_time, steps_per_interval)    
  
mydf_wk_end

#ggplot(mydf2, aes(datetime, avg_steps_day))+ geom_line(pch=8, col="blue")+ geom_hline(aes(yintercept= mean(avg_steps_day))) +
    #ggtitle("Average steps taken per day")


par(mfrow=c(2,1), mar=c(4,4,1,1))
plot(x=mydf_wk_end$pad_time, y=mydf_wk_end$steps_per_interval, main="Weekend" ,
     ylab="Number of Steps", xlab="Interval", type="l" ) + abline(h=mean(mydf_wk_end$steps_per_interval), col="blue")

plot(x=mydf_wk_day$pad_time, y=mydf_wk_day$steps_per_interval, main="Weekdays " ,ylab="Number of Steps", xlab="Interval", type="l" ) + abline(h=mean(mydf_wk_day$steps_per_interval), col="blue")




```

Findings:

Avg steps taken is more or less the same.  
